Model Recognition as Planning

### 2018-11-09 doc/icaps-19/model-recognition.tex
Compilation scheme explained

This section shows that $\delta^o(\mathcal{M},\mathcal{O})$, and hence an approximation to $P(\mathcal{O}|\mathcal{M})$), can be computed with a compilation-to-planning approach as the one proposed in~\cite{aineto2018learning} (AJO approach hereafter) for learning \strips\ models. The AJO approach receives as input an empty model $\mathcal{M}$, which only contains the headers of the action schemata formed of $\xi=\tup{name(\xi),pars(\xi)}$, and a set of observations of plan executions (i.e., a set of  $\mathcal{O}(\tau)$), and it returns a model $\mathcal{M'}$ with specification of preconditions and effects of each action schema included in $\mathcal{M}$ such that the validation of $\mathcal{O}(\tau)=\tup{s_0^o,a_1^o,s_1^o \ldots , a_l^o, s_m^o}$ following $\mathcal{M'}$ is successful; i.e., it holds $\rho(s_{i-1},a_i)$ for every observed action of $\mathcal{O}(\tau)$ and $s_i=\theta(s_{i-1},a_i)$ for every observed state of $\mathcal{O}(\tau)$.

In essence, the task $\tup{\mathcal{M},\mathcal{O}}$ of the AJO approach -- learning $\mathcal{M}$ that satisfies $\mathcal{O}$- can be interpreted as editing the empty action schemata of $\mathcal{M}$ introducing preconditions and effects until $\mathcal{O}$ is validated with the resulting model. We leverage the same idea to compute $\delta^o(\mathcal{M},\mathcal{O})$ with the exception that now our $\mathcal{M}$ is not empty but $\mathcal{M}=\{\xi_1, \ldots, \xi_m\}$, where $\xi_i=\tup{name(\xi_i),pars(\xi_i),pre(\xi_i),add(\xi_i),del(\xi_i)}$, $1 \leq i \leq m$.

Editing/learning the action schemata of $\mathcal{M}$ in the AJO approach is addressed converting the task into a classical planning problem, which is later solved with a planner. The intuition behind this compilation is that a solution plan to the problem is a sequence of: (a) \emph{edit actions} on the schemata of $\mathcal{M}$ to build $\mathcal{M'}$ and (b) \emph{validate actions} that apply $\mathcal{M'}$ in $\mathcal{O(\tau)}$. The adaptation of this compilation scheme for solving $\delta^o(\mathcal{M},\mathcal{O})$ results in a planning problem $P'=\tup{F',A',I',G'}$ whose objective is to determine the preconditions and effects of the action schemata of $\mathcal{M}$ that need to be added or deleted so as to satisfy $\mathcal{O}$. The accomplishment of this task requires therefore a propositional encoding of the components of the action schemata. Specifically:

\begin{itemize}
\item $F'$ contains the necessary fluents to represent the $pre$, $add$ and $del$ of the action schemata. Specifically, it is a set of \emph{editable fluents} of the type $\{pre\_e\_\xi, del\_e\_\xi, add\_e\_\xi\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}$ such that $e\in{\mathcal I}_{\Psi,\xi}$ is a single element from the set of FOL interpretations of predicates $\Psi$ over the corresponding parameters $pars(\xi)$. Additionally, $F$ also contains fluents to encode the observation constraints; that is, fluents to iterate through the $l$ observed actions and $m$ observed states of $\mathcal{O(\tau)}$.
\item $A'$ comprises two types of actions with conditional effects:
    \begin{itemize}
    \item actions for editing $\xi \in \mathcal{M}$ that follow the \emph{syntactic constraints} for well-defined \strips\ action schemata. Hence, here are actions for inserting a new precondition $pre\_e\_\xi$ or a new positive effect $add\_e\_\xi$ in $\xi$ when neither $pre\_e\_\xi$, $del\_e\_\xi$ or $add\_e\_\xi$ exist in $\xi$, and actions for inserting a new negative effect when neither $del\_e\_\xi$ or $add\_e\_\xi$ appear in $\xi$ but $pre\_e\_\xi$ does.
    \item actions for applying the new action schemata of the edited model $\mathcal{M'}$ and validating the observed states of $\mathcal{O(\tau)}$. Particularly, the \emph{apply} actions check the behavior of the propositional version of the new action schemata with the editable fluents of the current state.
    \end{itemize}
\item $I'$ encodes the editable fluents $pre\_e\_\xi, del\_e\_\xi$ and $add\_e\_\xi$ that hold in the action schemata of $\mathcal{M}$.
\item $G'$ contains the necessary fluents to check that all the observed actions and states of $\mathcal{O(\tau)}$ are generated correctly following the edited model $\mathcal{M'}$.
    %Here we include the necessary fluents to check the final reached state is consistent with the final state $s_m^o$ of $\mathcal{O(\tau)}$.
\end{itemize}

We now show an example of a solution plan to a planning problem $P'$ that results from compiling a specific task $\tup{\mathcal{M},\mathcal{O}}$. Consider that $\mathcal{M}$ is the model of Figure~\ref{fig:model-example} except that the schema {\tt\small inc-x} is defined without preconditions and its positive/negative effects are swapped wrt Figure~\ref{fig:model-example}. And consider an observation that only contains the initial and final state of Figure~\ref{fig:grid-example}; i.e., $\mathcal{O(\tau)}=\tup{s_0^o,s_m^o}=\tup{(1,1),(3,2)}$. The plan found by a planner to $P'$ is shown in figure~\ref{fig:plan-pdistance}. The first seven steps are the edit actions to \emph{fix} the schema {\tt\small inc-x}; step 07 is a validate action that sets the robot in the state $s_0^o=(1,1)$; step 08 applies the action {\tt\small (inc-x 1 2)} and so moves the robot one cell to the right to $(2,1)$; steps 09-11 move also the robot one cell to the right; step 11 applies an action that increases coordinate $y$ to an even number, thus moving robot to row $2$; step 12 moves the robot one cell to the left and, finally, action {\tt\small (validate\_1)} checks the robot position is consistent with the final state $s_m^o=(3,2)$.

>>>>

This section shows that $\delta^o(\mathcal{M},\mathcal{O})$, and hence an approximation to $P(\mathcal{O}|\mathcal{M})$), can be computed with a compilation-to-planning approach as the one proposed in~\cite{aineto2018learning} (AJO approach hereafter) for learning \strips\ models. The AJO approach receives as input an empty model $\mathcal{M}$, which only contains the headers of the action schemata formed of $\xi=\tup{name(\xi),pars(\xi)}$, and an observation of a plan execution $\mathcal{O}(\tau)$ (extensible to a set of observations), and it returns a model $\mathcal{M'}$ with specification of preconditions and effects of each action schema included in $\mathcal{M}$ such that the validation of $\mathcal{O}(\tau)=\tup{s_0^o,a_1^o,s_1^o \ldots , a_l^o, s_m^o}$ following $\mathcal{M'}$ is successful; i.e., it holds $\rho(s_{i-1},a_i)$ for every observed action of $\mathcal{O}(\tau)$ and $s_i=\theta(s_{i-1},a_i)$ for every observed state of $\mathcal{O}(\tau)$.

In essence, the task $\tup{\mathcal{M},\mathcal{O}}$ of the AJO approach -- learning $\mathcal{M}$ that satisfies $\mathcal{O}$-- can be interpreted as editing the empty action schemata of $\mathcal{M}$ introducing preconditions and effects until $\mathcal{O}$ is validated with the resulting model. We leverage the same idea to compute $\delta^o(\mathcal{M},\mathcal{O})$ with the exception that now our $\mathcal{M}$ is not empty but $\mathcal{M}=\{\xi_1, \ldots, \xi_m\}$, where $\xi_i=\tup{name(\xi_i),pars(\xi_i),pre(\xi_i),add(\xi_i),del(\xi_i)}$, $1 \leq i \leq m$.

Editing/learning the action schemata of $\mathcal{M}$ in the AJO approach is addressed converting the task into a classical planning problem, which is later solved with a planner. The intuition behind this compilation is that a solution plan to the problem is a sequence of: (a) \emph{edit actions} on the schemata of $\mathcal{M}$ to build $\mathcal{M'}$ and (b) \emph{validate actions} that apply $\mathcal{M'}$ in $\mathcal{O(\tau)}$. The adaptation of this compilation scheme for solving $\delta^o(\mathcal{M},\mathcal{O})$ results in a planning problem $P'=\tup{F',A',I',G'}$ whose objective is to determine the preconditions and effects that need to be added or deleted to the action schemata of $\mathcal{M}$ so as to satisfy $\mathcal{O}$. The accomplishment of this task requires therefore a propositional encoding of the components of the action schemata. Specifically:

\begin{itemize}
\item $F'$ contains the necessary fluents to represent the $pre$, $add$ and $del$ of the action schemata. It is a set of \emph{editable fluents} of the type $\{pre\_e\_\xi, del\_e\_\xi, add\_e\_\xi\}_{\forall e\in{\mathcal I}_{\Psi,\xi}}$ such that $e\in{\mathcal I}_{\Psi,\xi}$ is a single element from the set of FOL interpretations of predicates $\Psi$ over the corresponding parameters $pars(\xi)$. Additionally, $F$ also contains fluents to encode the \emph{observation constraints}; that is, fluents to iterate through the $l$ observed actions and $m$ observed states of $\mathcal{O(\tau)}$.
\item $A'$ comprises two types of actions with conditional effects:
    \begin{itemize}
    \item actions for editing $\xi \in \mathcal{M}$ that follow the \emph{syntactic constraints} for well-defined \strips\ action schemata. Hence, $A'$ contains actions for inserting a new precondition $pre\_e\_\xi$ or a new positive effect $add\_e\_\xi$ in $\xi$ when neither $pre\_e\_\xi$, $del\_e\_\xi$ or $add\_e\_\xi$ exist in $\xi$, and actions for inserting a new negative effect when neither $del\_e\_\xi$ or $add\_e\_\xi$ appear in $\xi$ but $pre\_e\_\xi$ does.
    \item actions for applying the new action schemata of the edited model $\mathcal{M'}$ and validating the observed states of $\mathcal{O(\tau)}$. Particularly, the \emph{apply} actions check the preconditions and produce the effects defined by the \emph{editable fluents}.
    \end{itemize}
\item $I'$ encodes the editable fluents $pre\_e\_\xi, del\_e\_\xi$ and $add\_e\_\xi$ that hold in the action schemata of $\mathcal{M}$.
\item $G'$ contains the necessary fluents to check that all the observed actions and states of $\mathcal{O(\tau)}$ are generated correctly following the edited model $\mathcal{M'}$.
    %Here we include the necessary fluents to check the final reached state is consistent with the final state $s_m^o$ of $\mathcal{O(\tau)}$.
\end{itemize}

We now show an example of a solution plan to a planning problem $P'$ that results from compiling a specific task $\tup{\mathcal{M},\mathcal{O}}$. Consider that $\mathcal{M}$ is the model of Figure~\ref{fig:model-example} except that the schema {\tt\small inc-x} is defined without preconditions and its positive/negative effects are swapped wrt Figure~\ref{fig:model-example}. And consider an observation that only contains the initial and final state of Figure~\ref{fig:grid-example}; i.e., $\mathcal{O(\tau)}=\tup{s_0^o,s_m^o}=\tup{(1,1),(3,2)}$. The plan found by a planner to $P'$ is shown in figure~\ref{fig:plan-pdistance}. The first seven steps are the edit actions to \emph{fix} the schema {\tt\small inc-x}; step 07 is a validate action that sets the robot in the initial state $s_0^o=(1,1)$; step 08 applies the action {\tt\small (inc-x 1 2)} and moves the robot one cell to the right to $(2,1)$; steps 09-11 also move the robot one cell to the right; step 11 applies an action that increases coordinate $y$ to an even number, thus moving robot to row $2$; step 12 moves the robot one cell to the left and, finally, action {\tt\small (validate\_1)} checks the robot position is consistent with the final state $s_m^o=(3,2)$.
